---
title: DuckDB & SQL
subtitle: "Lecture 21"
author: "Dr. Colin Rundel"
footer: "Sta 323 - Spring 2026"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    self-contained: true
execute:
  echo: true
  warning: true
engine: knitr
---


```{r setup}
#| message: False
#| warning: False
#| include: False
options(
  width=70
)

knitr::opts_chunk$set(
  fig.align = "center", fig.retina = 2, dpi = 150,
  out.width = "100%", warning = TRUE
)

library(dplyr)
```


## SQL

```{css}
/*| echo: false
code.sql > span > a::after {
  content: "D ";
  color: #aaaaaa;
}

pre.shell > code > span > a::after {
  content: "> ";
  color: #aaaaaa;
}
```

```{css}
/*| echo: false
.reveal pre code {
  text-wrap-mode: nowrap !important;
  max-height: 520px !important;
}
```

Structured Query Language is a special purpose language for interacting with (querying and modifying) indexed tabular data.

* ANSI Standard but with dialect divergence (MySql, Postgres, SQLite, etc.)

* This functionality maps very closely (but not exactly) with the data manipulation verbs present in dplyr.

* SQL is likely to be a foundational skill if you go into industry - learn it and put it on your CV


## DuckDB

> DuckDB is an open-source column-oriented relational database management system (RDBMS) originally developed by Mark Raasveldt and Hannes Mühleisen at the Centrum Wiskunde & Informatica (CWI) in the Netherlands and first released in 2019. The project has over 6 million downloads per month. It is designed to provide high performance on complex queries against large databases in embedded configuration, such as combining tables with hundreds of columns and billions of rows. Unlike other embedded databases (for example, SQLite) DuckDB is not focusing on transactional (OLTP) applications and instead is specialized for *online analytical processing (OLAP)* workloads.

From [Wikipedia - DuckDB](https://en.wikipedia.org/wiki/DuckDB)

## DuckDB & DBI

DuckDB is a relational database just like SQLite and can be interacted with using DBI and the duckdb package.

::: {.small}
```{r}
library(DBI)
(con = dbConnect(duckdb::duckdb()))
```
:::

. . .

::: {.small}
```{r}
dbWriteTable(con, "flights", nycflights13::flights)
dbListTables(con)
```
:::

##

::: {.small}
```{r}
dbGetQuery(con, "SELECT * FROM flights") |>
  as_tibble()
```
:::

##

::: {.small}
```{r}
#| message: false
library(dplyr)
tbl(con, "flights") |>
  filter(month == 10, day == 29) |>
  count(origin, dest) |>
  arrange(desc(n))
```
:::

# DuckDB CLI

## Connecting via CLI

::: {.small}
```shell
#| eval: false
duckdb employees.duckdb
```
```
DuckDB v1.5.1 (Variegata)
Enter ".help" for usage hints.
```
```sql
 
```
:::



## Table information

Dot commands are expressions that begins with `.` and are specific to the DuckDB CLI, some examples include:

::: {.small}
```sql
.tables
```
```
 ── employees ─── 
 ───── main ───── 
┌────────────────┐
│   employees    │
│                │
│ name   varchar │
│ email  varchar │
│ salary double  │
│ dept   varchar │
│                │
│     6 rows     │
└────────────────┘
```
:::

<p/>

::: {.small}
```sql
.schema employees
```
```
## CREATE TABLE employees("name" VARCHAR, email VARCHAR, salary DOUBLE, dept VARCHAR);
```
:::

<p/>

::: {.small}
```sql
.indexes employees
```

```sql
.maxrows 20
.maxwidth 80
```
:::

<br/>

A full list of available dot commands can be found [here](https://duckdb.org/docs/stable/clients/cli/dot_commands.html) or listed via `.help` in the CLI.


## SELECT Statements

```sql
SELECT * FROM employees;
```
```
## ┌─────────┬───────────────────┬─────────┬────────────┐
## │  name   │       email       │ salary  │    dept    │
## │ varchar │      varchar      │ double  │  varchar   │
## ├─────────┼───────────────────┼─────────┼────────────┤
## │ Alice   │ alice@company.com │ 52000.0 │ Accounting │
## │ Bob     │ bob@company.com   │ 40000.0 │ Accounting │
## │ Carol   │ carol@company.com │ 30000.0 │ Sales      │
## │ Dave    │ dave@company.com  │ 33000.0 │ Accounting │
## │ Eve     │ eve@company.com   │ 44000.0 │ Sales      │
## │ Frank   │ frank@company.com │ 37000.0 │ Sales      │
## └─────────┴───────────────────┴─────────┴────────────┘
```


## Output formats

The format of duckdb's output (in the CLI) is controlled via `.mode` -  the default is `duckbox`, see possible [output formats](https://duckdb.org/docs/stable/clients/cli/output_formats.html). 

:::: {.columns .xxsmall}
::: {.column width='50%'}
```sql
.mode csv
SELECT * FROM employees;
```
```
## name,email,salary,dept
## Alice,alice@company.com,52000.0,Accounting
## Bob,bob@company.com,40000.0,Accounting
## Carol,carol@company.com,30000.0,Sales
## Dave,dave@company.com,33000.0,Accounting
## Eve,eve@company.com,44000.0,Sales
## Frank,frank@company.com,37000.0,Sales
```
:::

::: {.column width='50%'}

```sql
.mode markdown
SELECT * FROM employees;
```
```
## | name  |       email       | salary  |    dept    |
## |-------|-------------------|--------:|------------|
## | Alice | alice@company.com | 52000.0 | Accounting |
## | Bob   | bob@company.com   | 40000.0 | Accounting |
## | Carol | carol@company.com | 30000.0 | Sales      |
## | Dave  | dave@company.com  | 33000.0 | Accounting |
## | Eve   | eve@company.com   | 44000.0 | Sales      |
## | Frank | frank@company.com | 37000.0 | Sales      |
```
:::
::::

:::: {.columns .xxsmall}
::: {.column width='50%'}
```sql
.mode json
SELECT * FROM employees;
```
```
## [{"name":"Alice","email":"alice@company.com","salary":52000.0,"dept":"Accounting"},
## {"name":"Bob","email":"bob@company.com","salary":40000.0,"dept":"Accounting"},
## {"name":"Carol","email":"carol@company.com","salary":30000.0,"dept":"Sales"},
## {"name":"Dave","email":"dave@company.com","salary":33000.0,"dept":"Accounting"},
## {"name":"Eve","email":"eve@company.com","salary":44000.0,"dept":"Sales"},
## {"name":"Frank","email":"frank@company.com","salary":37000.0,"dept":"Sales"}]
```
:::

::: {.column width='50%'}
```sql
.mode insert
SELECT * FROM employees;
```
```
INSERT INTO "table"("name",email,salary,dept) VALUES('Alice','alice@company.com',52000.0,'Accounting');
INSERT INTO "table"("name",email,salary,dept) VALUES('Bob','bob@company.com',40000.0,'Accounting');
INSERT INTO "table"("name",email,salary,dept) VALUES('Carol','carol@company.com',30000.0,'Sales');
INSERT INTO "table"("name",email,salary,dept) VALUES('Dave','dave@company.com',33000.0,'Accounting');
INSERT INTO "table"("name",email,salary,dept) VALUES('Eve','eve@company.com',44000.0,'Sales');
INSERT INTO "table"("name",email,salary,dept) VALUES('Frank','frank@company.com',37000.0,'Sales');
```
:::
::::


# A brief tour of SQL


## SELECT

We can subset and rename columns, sort results, and limit output using `SELECT`, `ORDER BY`, and `LIMIT`.

::: {.small}
```sql
SELECT name AS first_name, salary
FROM employees
ORDER BY salary DESC
LIMIT 3;
```
```
## ┌────────────┬─────────┐
## │ first_name │ salary  │
## │  varchar   │ double  │
## ├────────────┼─────────┤
## │ Alice      │ 52000.0 │
## │ Eve        │ 44000.0 │
## │ Bob        │ 40000.0 │
## └────────────┴─────────┘
```
:::

. . .

`DISTINCT` removes duplicate values:

::: {.small}
```sql
SELECT DISTINCT dept FROM employees;
```
```
## ┌────────────┐
## │    dept    │
## │  varchar   │
## ├────────────┤
## │ Accounting │
## │ Sales      │
## └────────────┘
```
:::

::: {.aside}
SQL clauses are evaluated in a different order than they are written: `FROM` → `WHERE` → `GROUP BY` → `SELECT` → `ORDER BY` → `LIMIT`
:::


## filter() using WHERE

We can filter rows using a `WHERE` clause

::: {.small}
```sql
SELECT * FROM employees WHERE salary < 40000;
```
```
## ┌─────────┬───────────────────┬─────────┬────────────┐
## │  name   │       email       │ salary  │    dept    │
## │ varchar │      varchar      │ double  │  varchar   │
## ├─────────┼───────────────────┼─────────┼────────────┤
## │ Carol   │ carol@company.com │ 30000.0 │ Sales      │
## │ Dave    │ dave@company.com  │ 33000.0 │ Accounting │
## │ Frank   │ frank@company.com │ 37000.0 │ Sales      │
## └─────────┴───────────────────┴─────────┴────────────┘
```

```sql
SELECT * FROM employees WHERE salary < 40000 AND dept = 'Sales';
```
```
## ┌─────────┬───────────────────┬─────────┬─────────┐
## │  name   │       email       │ salary  │  dept   │
## │ varchar │      varchar      │ double  │ varchar │
## ├─────────┼───────────────────┼─────────┼─────────┤
## │ Carol   │ carol@company.com │ 30000.0 │ Sales   │
## │ Frank   │ frank@company.com │ 37000.0 │ Sales   │
## └─────────┴───────────────────┴─────────┴─────────┘
```
:::

## GROUP BY and HAVING

`GROUP BY` groups rows for aggregation; `HAVING` filters *after* aggregation (like `WHERE` but for grouped results).

::: {.small}
```sql
SELECT dept, COUNT(*) AS n, round(AVG(salary),2) AS avg_salary
FROM employees GROUP BY dept;
```
```
## ┌────────────┬───────┬────────────┐
## │    dept    │   n   │ avg_salary │
## │  varchar   │ int64 │   double   │
## ├────────────┼───────┼────────────┤
## │ Sales      │     3 │    37000.0 │
## │ Accounting │     3 │  41666.67  │
## └────────────┴───────┴────────────┘
```
:::

. . .

::: {.small}
```sql
SELECT dept, COUNT(*) AS n, round(AVG(salary),2) AS avg_salary
FROM employees GROUP BY dept
HAVING avg_salary > 38000;
```
```
## ┌────────────┬───────┬────────────┐
## │    dept    │   n   │ avg_salary │
## │  varchar   │ int64 │   double   │
## ├────────────┼───────┼────────────┤
## │ Accounting │     3 │  41666.67  │
## └────────────┴───────┴────────────┘
```
:::



## Exercise 1

Using duckdb calculate the following quantities for `employees.duckdb`,

1. The total costs in payroll for this company

2. The average salary within each department

::: {.aside}
[DuckDB's aggregation functions](https://duckdb.org/docs/stable/sql/functions/aggregates.html)
:::


## Reading from CSV files

DuckDB has a neat trick in that it can treat files as tables (for supported formats), this lets you query them without having to explicitly create a table in the database.

We can also make this explicit by using the `read_csv()` function, which is useful if we need to use custom options (e.g. specify a different delimiter)

:::: {.columns .xsmall}
::: {.column width='50%'}
```sql
SELECT * FROM 'phone.csv';
```
```
## ┌─────────┬──────────────┐
## │  name   │    phone     │
## │ varchar │   varchar    │
## ├─────────┼──────────────┤
## │ Bob     │ 919 555-1111 │
## │ Carol   │ 919 555-2222 │
## │ Eve     │ 919 555-3333 │
## │ Frank   │ 919 555-4444 │
## └─────────┴──────────────┘
```
:::

::: {.column width='50%'}
```sql
SELECT * FROM 
  read_csv('phone.csv', delim = ',');
```
```
## ┌─────────┬──────────────┐
## │  name   │    phone     │
## │ varchar │   varchar    │
## ├─────────┼──────────────┤
## │ Bob     │ 919 555-1111 │
## │ Carol   │ 919 555-2222 │
## │ Eve     │ 919 555-3333 │
## │ Frank   │ 919 555-4444 │
## └─────────┴──────────────┘
```
:::
::::

## Tables from CSV

If we wanted to explicitly create a table from the CSV file this is also possible, 

::: {.small .extra-pb}
```sql
.tables
```
```
 ── employees ─── 
 ───── main ───── 
┌────────────────┐
│   employees    │
│                │
│ name   varchar │
│ email  varchar │
│ salary double  │
│ dept   varchar │
│                │
│     6 rows     │
└────────────────┘
```
:::

. . .

::: {.small .extra-pb}
```sql
CREATE TABLE phone AS
  SELECT * FROM 'phone.csv';
```
```sql
.tables
```
```
 ─────────── employees ─────────── 
 ───────────── main ────────────── 
┌────────────────┐┌───────────────┐
│   employees    ││     phone     │
│                ││               │
│ name   varchar ││ name  varchar │
│ email  varchar ││ phone varchar │
│ salary double  ││               │
│ dept   varchar ││    4 rows     │
│                │└───────────────┘
│     6 rows     │
└────────────────┘
```
:::

. . .

::: {.small}
```sql
SELECT * FROM phone;
```
```
## ┌─────────┬──────────────┐
## │  name   │    phone     │
## │ varchar │   varchar    │
## ├─────────┼──────────────┤
## │ Bob     │ 919 555-1111 │
## │ Carol   │ 919 555-2222 │
## │ Eve     │ 919 555-3333 │
## │ Frank   │ 919 555-4444 │
## └─────────┴──────────────┘
```
:::

## Views from CSV

It is also possible to create a view from a file - this acts like a table but the data is not copied from the file. This means the file must remain accessible and is re-read on every query, but no storage is used within the database.

::: {.small .extra-pb}
```sql
.tables
```
```
 ─────────── employees ─────────── 
 ───────────── main ────────────── 
┌────────────────┐┌───────────────┐
│   employees    ││     phone     │
│                ││               │
│ name   varchar ││ name  varchar │
│ email  varchar ││ phone varchar │
│ salary double  ││               │
│ dept   varchar ││    4 rows     │
│                │└───────────────┘
│     6 rows     │
└────────────────┘
```
:::

. . .

::: {.small .extra-pb}
```sql
CREATE VIEW phone_view AS
  SELECT * FROM 'phone.csv';
```
```sql
.tables
```
```
 ─────────────────── employees ──────────────────── 
 ────────────────────── main ────────────────────── 
┌────────────────┐┌───────────────┐┌───────────────┐
│   employees    ││     phone     ││  phone_view   │
│                ││               ││               │
│ name   varchar ││ name  varchar ││ name  varchar │
│ email  varchar ││ phone varchar ││ phone varchar │
│ salary double  ││               │└───────────────┘
│ dept   varchar ││    4 rows     │
│                │└───────────────┘
│     6 rows     │                 
└────────────────┘  
```
:::

. . .

::: {.small .extra-pb}
```sql
SELECT * FROM phone_view;
```
```
## ┌─────────┬──────────────┐
## │  name   │    phone     │
## │ varchar │   varchar    │
## ├─────────┼──────────────┤
## │ Bob     │ 919 555-1111 │
## │ Carol   │ 919 555-2222 │
## │ Eve     │ 919 555-3333 │
## │ Frank   │ 919 555-4444 │
## └─────────┴──────────────┘
```
:::

## Parquet

Parquet is a binary, columnar file format widely used in data science and analytics.

:::: {.columns}
::: {.column width='50%'}
**Row-oriented (CSV)**

```
name,  salary, dept
Alice, 52000,  Accounting
Bob,   40000,  Accounting
Carol, 30000,  Sales
```
Reading one column requires reading every row.
:::

::: {.column width='50%'}
**Column-oriented (Parquet)**

```
name:   [Alice, Bob, Carol, ...]
salary: [52000, 40000, 30000, ...]
dept:   [Accounting, Accounting, Sales, ...]
```
Reading one column skips all others entirely.
:::
::::

Key advantages over CSV: **compressed** (much smaller files), **typed** (no schema inference), **fast for analytical queries** (only reads needed columns). Widely used with Spark, Arrow, Pandas, and cloud data warehouses.


## Reading from Parquet files

DuckDB can query Parquet files directly — no explicit import step needed.

:::: {.columns .small}
::: {.column width='50%'}
```sql
SELECT * FROM 'data.parquet' LIMIT 5;
```
:::
::: {.column width='50%'}
```sql
SELECT * FROM 'data/*.parquet' LIMIT 5;
```
:::
::::

Glob patterns let you query multiple files as a single table — useful for partitioned datasets.

. . .

The same `CREATE TABLE` / `CREATE VIEW` pattern from CSV works with Parquet:

::: {.small}
```sql
CREATE VIEW flights AS SELECT * FROM 'flights/*.parquet';
```
:::

::: {.aside}
DuckDB can also read remote Parquet files directly via HTTPS (with the `httpfs` extension), enabling queries over S3 or other cloud storage without downloading the file first.
:::


## Deleting tables and views

Tables and views can be deleted using `DROP`

::: {.small .extra-pb}

```sql
.tables
```
```
## employees  phone  phone_view
```

```sql
DROP TABLE phone;
DROP VIEW phone_view;
```
:::

. . .

::: {.small .extra-pb}

```sql
.tables
```
```
## employees
```
:::


## Inner Join

DuckDB requires an `ON` or `USING` clause for joins — `JOIN` alone is a syntax error.

::: {.small}
```sql
SELECT * FROM employees JOIN phone USING(name);
```
```
## ┌─────────┬───────────────────┬─────────┬────────────┬──────────────┐
## │  name   │       email       │ salary  │    dept    │    phone     │
## │ varchar │      varchar      │ double  │  varchar   │   varchar    │
## ├─────────┼───────────────────┼─────────┼────────────┼──────────────┤
## │ Bob     │ bob@company.com   │ 40000.0 │ Accounting │ 919 555-1111 │
## │ Carol   │ carol@company.com │ 30000.0 │ Sales      │ 919 555-2222 │
## │ Eve     │ eve@company.com   │ 44000.0 │ Sales      │ 919 555-3333 │
## │ Frank   │ frank@company.com │ 37000.0 │ Sales      │ 919 555-4444 │
## └─────────┴───────────────────┴─────────┴────────────┴──────────────┘
```
:::

::: {.aside}
`USING(col)` joins on a named column (no duplicate column in result); `ON t1.col = t2.col` is needed when column names differ. `NATURAL JOIN` automatically matches all common column names — convenient but can produce unexpected results when tables share more columns than intended (more on this shortly).
:::

## Left Join

`LEFT JOIN` keeps all rows from the left table, filling `NULL` for unmatched rows in the right table.

::: {.small}
```sql
SELECT * FROM employees LEFT JOIN phone USING(name);
```
```
## ┌─────────┬───────────────────┬─────────┬────────────┬──────────────┐
## │  name   │       email       │ salary  │    dept    │    phone     │
## │ varchar │      varchar      │ double  │  varchar   │   varchar    │
## ├─────────┼───────────────────┼─────────┼────────────┼──────────────┤
## │ Bob     │ bob@company.com   │ 40000.0 │ Accounting │ 919 555-1111 │
## │ Carol   │ carol@company.com │ 30000.0 │ Sales      │ 919 555-2222 │
## │ Eve     │ eve@company.com   │ 44000.0 │ Sales      │ 919 555-3333 │
## │ Frank   │ frank@company.com │ 37000.0 │ Sales      │ 919 555-4444 │
## │ Alice   │ alice@company.com │ 52000.0 │ Accounting │              │
## │ Dave    │ dave@company.com  │ 33000.0 │ Accounting │              │
## └─────────┴───────────────────┴─────────┴────────────┴──────────────┘
```
:::

## Other Joins

All other standard joins are supported: `RIGHT JOIN`, `FULL JOIN`, `CROSS JOIN`, `SEMI JOIN`, `ANTI JOIN`, etc.

::: {.small}
```sql
SELECT * FROM employees FULL JOIN phone USING(name);
SELECT * FROM employees RIGHT JOIN phone USING(name);
SELECT * FROM employees ANTI JOIN phone USING(name);
```
:::

## Subqueries

Tables can be nested within tables for the purpose of queries,

::: {.xsmall}
```sql
SELECT * FROM (
  SELECT * FROM employees NATURAL LEFT JOIN phone
) combined WHERE phone IS NULL;
```
```
## ┌─────────┬───────────────────┬─────────┬────────────┬─────────┐
## │  name   │       email       │ salary  │    dept    │  phone  │
## │ varchar │      varchar      │ double  │  varchar   │ varchar │
## ├─────────┼───────────────────┼─────────┼────────────┼─────────┤
## │ Alice   │ alice@company.com │ 52000.0 │ Accounting │         │
## │ Dave    │ dave@company.com  │ 33000.0 │ Accounting │         │
## └─────────┴───────────────────┴─────────┴────────────┴─────────┘
```

```sql
SELECT * FROM (
  SELECT * FROM employees NATURAL LEFT JOIN phone
) combined WHERE phone IS NOT NULL;
```
```
## ┌─────────┬───────────────────┬─────────┬────────────┬──────────────┐
## │  name   │       email       │ salary  │    dept    │    phone     │
## │ varchar │      varchar      │ double  │  varchar   │   varchar    │
## ├─────────┼───────────────────┼─────────┼────────────┼──────────────┤
## │ Bob     │ bob@company.com   │ 40000.0 │ Accounting │ 919 555-1111 │
## │ Carol   │ carol@company.com │ 30000.0 │ Sales      │ 919 555-2222 │
## │ Eve     │ eve@company.com   │ 44000.0 │ Sales      │ 919 555-3333 │
## │ Frank   │ frank@company.com  │ 37000.0 │ Sales      │ 919 555-4444 │
## └─────────┴───────────────────┴─────────┴────────────┴──────────────┘
```
:::


## Exercise 2

Let's try to create a table that has a new column - `abv_avg` which contains how much more (or less) than the average, for their department, each person is paid.

Hint - This will require joining a subquery.

<!--
```sql
SELECT *, round(salary-avg,2) AS abv_avg
FROM employees
NATURAL JOIN  (
  SELECT dept, round(avg(salary),2) AS avg FROM employees GROUP BY dept
) dept_avg
ORDER BY dept, abv_avg;
## ┌─────────┬───────────────────┬─────────┬────────────┬──────────┬─────────┐
## │  name   │       email       │ salary  │    dept    │   avg    │ abv_avg │
## │ varchar │      varchar      │ double  │  varchar   │  double  │ double  │
## ├─────────┼───────────────────┼─────────┼────────────┼──────────┼─────────┤
## │ Alice   │ alice@company.com │ 52000.0 │ Accounting │ 41666.67 │ 10333.0 │
## │ Bob     │ bob@company.com   │ 40000.0 │ Accounting │ 41666.67 │ -1667.0 │
## │ Carol   │ carol@company.com │ 30000.0 │ Sales      │  37000.0 │ -7000.0 │
## │ Dave    │ dave@company.com  │ 33000.0 │ Accounting │ 41666.67 │ -8667.0 │
## │ Eve     │ eve@company.com   │ 44000.0 │ Sales      │  37000.0 │  7000.0 │
## │ Frank   │ frank@company.com │ 37000.0 │ Sales      │  37000.0 │     0.0 │
## └─────────┴───────────────────┴─────────┴────────────┴──────────┴─────────┘
```
-->


# Query plan

## Setup

To give us a bit more variety (and data), we have created another DuckDB database `flights.duckdb` that contains both `nycflights13::flights` and `nycflights13::planes`, the latter of which has details on each plane identified by tail number.

::: {.small}
```{r}
#| eval: False
db = DBI::dbConnect(duckdb::duckdb(), "flights.duckdb")
dplyr::copy_to(db, nycflights13::flights, name = "flights", temporary = FALSE, overwrite = TRUE)
dplyr::copy_to(db, nycflights13::planes, name = "planes", temporary = FALSE, overwrite = TRUE)
DBI::dbDisconnect(db)
```
:::

. . .

All of the following code will be run in the DuckDB command line interface, make sure you've created the database and copied both the flights and planes tables into the db or use the version provided in the `exercises/` repo.


## Opening `flights.duckdb`

The database can then be opened from the terminal tab using,
```shell
#| eval: false
duckdb flights.duckdb
```

As before we set a couple of configuration options so that our output is readable. We also include `.timer on` so that we get timings for our queries.

```sql
.maxrows 20
.maxwidth 80
.timer on
```

## `flights`

::: {.small}
```sql
SELECT * FROM flights LIMIT 10;
```
```
## ┌───────┬───────┬───────┬───┬──────────┬────────┬────────┬─────────────────────┐
## │ year  │ month │  day  │ … │ distance │  hour  │ minute │      time_hour      │
## │ int32 │ int32 │ int32 │   │  double  │ double │ double │      timestamp      │
## ├───────┼───────┼───────┼───┼──────────┼────────┼────────┼─────────────────────┤
## │  2013 │     1 │     1 │ … │   1400.0 │    5.0 │   15.0 │ 2013-01-01 10:00:00 │
## │  2013 │     1 │     1 │ … │   1416.0 │    5.0 │   29.0 │ 2013-01-01 10:00:00 │
## │  2013 │     1 │     1 │ … │   1089.0 │    5.0 │   40.0 │ 2013-01-01 10:00:00 │
## │  2013 │     1 │     1 │ … │   1576.0 │    5.0 │   45.0 │ 2013-01-01 10:00:00 │
## │  2013 │     1 │     1 │ … │    762.0 │    6.0 │    0.0 │ 2013-01-01 11:00:00 │
## │  2013 │     1 │     1 │ … │    719.0 │    5.0 │   58.0 │ 2013-01-01 10:00:00 │
## │  2013 │     1 │     1 │ … │   1065.0 │    6.0 │    0.0 │ 2013-01-01 11:00:00 │
## │  2013 │     1 │     1 │ … │    229.0 │    6.0 │    0.0 │ 2013-01-01 11:00:00 │
## │  2013 │     1 │     1 │ … │    944.0 │    6.0 │    0.0 │ 2013-01-01 11:00:00 │
## │  2013 │     1 │     1 │ … │    733.0 │    6.0 │    0.0 │ 2013-01-01 11:00:00 │
## ├───────┴───────┴───────┴───┴──────────┴────────┴────────┴─────────────────────┤
## │ 10 rows                                                 19 columns (7 shown) │
## └──────────────────────────────────────────────────────────────────────────────┘
## Run Time (s): real 0.020 user 0.000784 sys 0.002284
```
:::

## `planes`

::: {.small}
```sql
SELECT * FROM planes LIMIT 10;
```
```
## ┌─────────┬───────┬──────────────────────┬───┬───────┬───────┬───────────┐
## │ tailnum │ year  │         type         │ … │ seats │ speed │  engine   │
## │ varchar │ int32 │       varchar        │   │ int32 │ int32 │  varchar  │
## ├─────────┼───────┼──────────────────────┼───┼───────┼───────┼───────────┤
## │ N10156  │  2004 │ Fixed wing multi e…  │ … │    55 │       │ Turbo-fan │
## │ N102UW  │  1998 │ Fixed wing multi e…  │ … │   182 │       │ Turbo-fan │
## │ N103US  │  1999 │ Fixed wing multi e…  │ … │   182 │       │ Turbo-fan │
## │ N104UW  │  1999 │ Fixed wing multi e…  │ … │   182 │       │ Turbo-fan │
## │ N10575  │  2002 │ Fixed wing multi e…  │ … │    55 │       │ Turbo-fan │
## │ N105UW  │  1999 │ Fixed wing multi e…  │ … │   182 │       │ Turbo-fan │
## │ N107US  │  1999 │ Fixed wing multi e…  │ … │   182 │       │ Turbo-fan │
## │ N108UW  │  1999 │ Fixed wing multi e…  │ … │   182 │       │ Turbo-fan │
## │ N109UW  │  1999 │ Fixed wing multi e…  │ … │   182 │       │ Turbo-fan │
## │ N110UW  │  1999 │ Fixed wing multi e…  │ … │   182 │       │ Turbo-fan │
## ├─────────┴───────┴──────────────────────┴───┴───────┴───────┴───────────┤
## │ 10 rows                                            9 columns (6 shown) │
## └────────────────────────────────────────────────────────────────────────┘
## Run Time (s): real 0.003 user 0.000819 sys 0.000018
```
:::


## Exercise 3

Write a query that determines the total number of seats available on all of the planes that flew out of New York in 2013.


## A Solution?

Does the following seem correct?

::: {.small}
```sql
SELECT sum(seats) FROM flights NATURAL LEFT JOIN planes;
```
```
## ┌────────────┐
## │ sum(seats) │
## │   int128   │
## ├────────────┤
## │     614366 │
## └────────────┘
## Run Time (s): real 0.012 user 0.016061 sys 0.002386
```
:::

. . .

<br/><br/>

::: {.center .large}
Why?
:::


## Correct solution

**Join and select**:

::: {.small}
```sql
SELECT sum(seats) FROM flights LEFT JOIN planes USING (tailnum);
```
```
## ┌────────────┐
## │ sum(seats) │
## │   int128   │
## ├────────────┤
## │   38851317 │
## └────────────┘
## Run Time (s): real 0.005 user 0.010150 sys 0.000291
```
:::



## `EXPLAIN`

::: {.xxsmall}
```sql
EXPLAIN SELECT sum(seats) FROM flights LEFT JOIN planes USING (tailnum);
```
```
## ┌─────────────────────────────┐
## │┌───────────────────────────┐│
## ││       Physical Plan       ││
## │└───────────────────────────┘│
## └─────────────────────────────┘
## ┌───────────────────────────┐
## │    UNGROUPED_AGGREGATE    │
## │    ────────────────────   │
## │    Aggregates: sum(#0)    │
## └─────────────┬─────────────┘
## ┌─────────────┴─────────────┐
## │         PROJECTION        │
## │    ────────────────────   │
## │           seats           │
## │                           │
## │        ~336776 Rows       │
## └─────────────┬─────────────┘
## ┌─────────────┴─────────────┐
## │         HASH_JOIN         │
## │    ────────────────────   │
## │      Join Type: LEFT      │
## │                           │
## │        Conditions:        ├──────────────┐
## │     tailnum = tailnum     │              │
## │                           │              │
## │        ~336776 Rows       │              │
## └─────────────┬─────────────┘              │
## ┌─────────────┴─────────────┐┌─────────────┴─────────────┐
## │         SEQ_SCAN          ││         SEQ_SCAN          │
## │    ────────────────────   ││    ────────────────────   │
## │          flights          ││           planes          │
## │                           ││                           │
## │    Projections: tailnum   ││        Projections:       │
## │                           ││          tailnum          │
## │                           ││           seats           │
## │                           ││                           │
## │        ~336776 Rows       ││         ~3322 Rows        │
## └───────────────────────────┘└───────────────────────────┘
## Run Time (s): real 0.001 user 0.000547 sys 0.000000
```
:::

## `EXPLAIN ANALYZE`

::: {.xxsmall}
```sql
EXPLAIN ANALYZE SELECT sum(seats) FROM flights LEFT JOIN planes USING (tailnum);
```
```
## ┌─────────────────────────────────────┐
## │┌───────────────────────────────────┐│
## ││    Query Profiling Information    ││
## │└───────────────────────────────────┘│
## └─────────────────────────────────────┘
## EXPLAIN ANALYZE SELECT sum(seats) FROM flights LEFT JOIN planes USING (tailnum);
## ┌────────────────────────────────────────────────┐
## │┌──────────────────────────────────────────────┐│
## ││              Total Time: 0.0045s             ││
## │└──────────────────────────────────────────────┘│
## └────────────────────────────────────────────────┘
## ┌───────────────────────────┐
## │           QUERY           │
## └─────────────┬─────────────┘
## ┌─────────────┴─────────────┐
## │      EXPLAIN_ANALYZE      │
## │    ────────────────────   │
## │           0 Rows          │
## │          (0.00s)          │
## └─────────────┬─────────────┘
## ┌─────────────┴─────────────┐
## │    UNGROUPED_AGGREGATE    │
## │    ────────────────────   │
## │    Aggregates: sum(#0)    │
## │                           │
## │           1 Rows          │
## │          (0.00s)          │
## └─────────────┬─────────────┘
## ┌─────────────┴─────────────┐
## │         PROJECTION        │
## │    ────────────────────   │
## │           seats           │
## │                           │
## │        336776 Rows        │
## │          (0.00s)          │
## └─────────────┬─────────────┘
## ┌─────────────┴─────────────┐
## │         HASH_JOIN         │
## │    ────────────────────   │
## │      Join Type: LEFT      │
## │                           │
## │        Conditions:        ├──────────────┐
## │     tailnum = tailnum     │              │
## │                           │              │
## │        336776 Rows        │              │
## │          (0.01s)          │              │
## └─────────────┬─────────────┘              │
## ┌─────────────┴─────────────┐┌─────────────┴─────────────┐
## │         TABLE_SCAN        ││         TABLE_SCAN        │
## │    ────────────────────   ││    ────────────────────   │
## │          flights          ││           planes          │
## │                           ││                           │
## │    Projections: tailnum   ││        Projections:       │
## │                           ││          tailnum          │
## │                           ││           seats           │
## │                           ││                           │
## │        336776 Rows        ││         3322 Rows         │
## │          (0.00s)          ││          (0.00s)          │
## └───────────────────────────┘└───────────────────────────┘
## Run Time (s): real 0.004 user 0.011027 sys 0.000200
```
:::

## dplyr

::: {.small}
```{r}
library(dplyr)
flights = nycflights13::flights
planes = nycflights13::planes

system.time({
  flights |>
    left_join(planes, by = c("tailnum" = "tailnum")) |>
    summarise(total_seats = sum(seats, na.rm = TRUE))
})
```
:::


# NYC Taxi Demo