---
title: "Optimization (cont.)"
subtitle: "Lecture 25"
author: "Dr. Colin Rundel"
footer: "Sta 323 - Spring 2026"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    self-contained: true
execute:
  echo: true
engine: knitr
---

```{r setup}
#| message: false
#| warning: false
#| include: false

knitr::opts_chunk$set(
  fig.align = "center", fig.retina = 2, dpi = 150,
  out.width = "100%"
)

library(tidyverse)
ggplot2::theme_set(ggplot2::theme_bw())

options(width = 70)
```


# Regression via `optim()`

## Simulated data

```{r}
set.seed(42)
n = 200
d = tibble(
  x1 = rnorm(n),
  x2 = rnorm(n),
  x3 = rnorm(n),
  x4 = rnorm(n),
  x5 = rnorm(n),
  y = 3 + 2*x1 - 1.5*x2 + 0*x3 + 0.1*x4 + 0*x5 + rnorm(n)
)
```

True coefficients: $\beta_0 = 3$, $\beta_1 = 2$, $\beta_2 = -1.5$, $\beta_3 = 0$, $\beta_4 = 0.1$, $\beta_5 = 0$


## OLS as optimization

The least squares objective is

$$\text{RSS}(\beta) = \sum_{i=1}^n \left(y_i - \beta_0 - \sum_{j=1}^p \beta_j x_{ij}\right)^2$$

::: {.small}
```{r}
X = model.matrix(y ~ ., data = d)

rss = function(beta) {
  sum((d$y - X %*% beta)^2)
}
```
:::


## Fitting with `optim()`

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
res_ols = optim(
  rep(0, 6), rss, method = "BFGS"
)
res_ols
```
:::

::: {.column width='50%' .fragment}
```{r}
coef(lm(y ~ ., data = d))
```
:::
::::


## Lasso regression

The Lasso adds an $L_1$ penalty to shrink coefficients toward zero (note: we don't penalize the intercept),

$$\text{RSS}(\beta) + \lambda \sum_{j=1}^p |\beta_j|$$

. . .

::: {.small}
```{r}
lasso_obj = function(beta, lambda) {
  sum((d$y - X %*% beta)^2) + lambda * sum(abs(beta[-1]))
}
```
:::


## Lasso results

Since the $L_1$ penalty is not differentiable, we use Nelder-Mead.

::: {.xsmall}
```{r}
fit_lasso = function(lambda) {
  res = optim(rep(0, 6), lasso_obj, lambda = lambda, method = "Nelder-Mead",
              control = list(maxit = 10000))
  setNames(res$par, colnames(X))
}
```
:::

. . .

::: {.xsmall}
```{r}
rbind(
  "OLS"           = coef(lm(y ~ ., data = d)),
  "Lasso (λ=10)"  = fit_lasso(10),
  "Lasso (λ=50)"  = fit_lasso(50),
  "Lasso (λ=100)" = fit_lasso(100),
  "Truth"         = c(3, 2, -1.5, 0, 0.1, 0)
) |> round(3)
```
:::


# Maximum Likelihood Estimation

## The idea

Given observed data $x_1, \ldots, x_n$ and a parametric model $f(x | \theta)$, the **likelihood** is

$$\mathcal{L}(\theta | \mathbf{x}) = \prod_{i=1}^n f(x_i | \theta)$$

. . .

The **log-likelihood** is easier to work with,

$$\ell(\theta | \mathbf{x}) = \sum_{i=1}^n \log f(x_i | \theta)$$

. . .

The **MLE** is the parameter value that maximizes the likelihood (equivalently, minimizes the negative log-likelihood),

$$\hat{\theta}_{\text{MLE}} = \arg\max_\theta \, \ell(\theta | \mathbf{x}) = \arg\min_\theta \, -\ell(\theta | \mathbf{x})$$


## Normal MLE

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
set.seed(1234)
x = rnorm(100, mean = -3.2, sd = 1.25)
c(mean = mean(x), sd = sd(x))
```
```{r}
nll_norm = function(theta) {
  -sum(dnorm(x, mean = theta[1],
             sd = theta[2], log = TRUE))
}
```
:::

::: {.column width='50%' .fragment}
```{r}
nll_norm(c(0, 1))
nll_norm(c(-3, 1))
nll_norm(c(-3.2, 1.25))
```
:::
::::


## Fitting the Normal MLE

::: {.small}
```{r}
optim(c(0, 1), nll_norm, method = "BFGS")
```
:::

. . .

Note: `sd` must be positive, but BFGS doesn't know that — we got lucky with our starting value. We'll address this shortly.


## Gamma MLE

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
set.seed(5678)
x_gam = rgamma(200, shape = 3, rate = 2)
c(mean = mean(x_gam), var = var(x_gam))
```
```{r}
nll_gamma = function(theta) {
  -sum(dgamma(x_gam, shape = theta[1],
              rate = theta[2], log = TRUE))
}
```
:::

::: {.column width='50%' .fragment}
```{r}
optim(
  c(1, 1), nll_gamma, method = "BFGS"
)
```
:::
::::


## Beta MLE

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
set.seed(9012)
x_beta = rbeta(200, shape1 = 2, shape2 = 5)
c(mean = mean(x_beta), var = var(x_beta))
```
```{r}
nll_beta = function(theta) {
  -sum(dbeta(x_beta, shape1 = theta[1],
             shape2 = theta[2], log = TRUE))
}
```
:::

::: {.column width='50%' .fragment}
```{r}
optim(
  c(1, 1), nll_beta, method = "BFGS"
)
```
:::
::::


# Constrained Parameters

## The problem

Many distribution parameters have restricted support — e.g., $\sigma > 0$, shape $> 0$, rate $> 0$.

If `optim()` tries a negative value, the log-likelihood returns `NaN`:

::: {.small}
```{r}
dnorm(0, mean = 0, sd = -1, log = TRUE)
dgamma(1, shape = -1, rate = 1, log = TRUE)
```
:::

. . .

This causes the optimizer to fail or give unreliable results. Two common solutions:

1. **Reparameterization** — optimize over a transformed, unconstrained space
2. **Box constraints** — use `L-BFGS-B` with `lower` / `upper` bounds


## Reparameterization

Optimize over $\log(\sigma)$ instead of $\sigma$, using $\exp()$ to map back to the positive reals.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
nll_norm_reparam = function(theta) {
  -sum(dnorm(
    x, mean = theta[1],
    sd = exp(theta[2]), log = TRUE
  ))
}
```
:::

::: {.column width='50%' .fragment}
```{r}
res = optim(
  c(0, log(1)), nll_norm_reparam,
  method = "BFGS"
)
res
```
:::
::::

. . .

::: {.xsmall}
```{r}
c(mu = res$par[1], sigma = exp(res$par[2]))
```
:::


## Reparameterized Gamma MLE

Both shape and rate must be positive — optimize over their logs.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
nll_gamma_reparam = function(theta) {
  -sum(dgamma(
    x_gam,
    shape = exp(theta[1]),
    rate = exp(theta[2]),
    log = TRUE
  ))
}
```
:::

::: {.column width='50%' .fragment}
```{r}
res_gam = optim(
  c(0, 0), nll_gamma_reparam,
  method = "BFGS"
)
exp(res_gam$par)
```
:::
::::


## Box constraints with `L-BFGS-B`

An alternative — specify bounds directly via `lower` and `upper` (only available for `"L-BFGS-B"`).

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
optim(
  c(0, 1), nll_norm,
  method = "L-BFGS-B",
  lower = c(-Inf, 1e-6),
  upper = c(Inf, Inf)
)
```
:::

::: {.column width='50%' .fragment}
```{r}
optim(
  c(1, 1), nll_gamma,
  method = "L-BFGS-B",
  lower = c(1e-6, 1e-6),
  upper = c(Inf, Inf)
)
```
:::
::::


## Reparameterization vs. bounds

| | Reparameterization | Box Constraints |
|---|---|---|
| **Methods** | Any (`BFGS`, `CG`, ...) | `L-BFGS-B` only |
| **Gradients** | Need chain rule adjustment | Work on original scale |
| **Flexibility** | Any smooth transform | Lower/upper bounds only |
| **Edge behavior** | Can't reach boundary exactly | Can sit on boundary |

. . .

In general, **reparameterization is more flexible** and works with any method. Use **bounds** when you want simplicity and `L-BFGS-B` is sufficient.


# Automatic Differentiation in Base R

## The gradient problem

Throughout this lecture we've been supplying hand-coded gradient functions to `optim()`. This works but is:

* **Error-prone** — easy to make mistakes in the calculus or the code

* **Tedious** — especially for complex objective functions

* **Hard to maintain** — changing the objective means updating the gradient too

. . .

Base R provides tools for **symbolic differentiation** that can compute derivatives automatically: `D()`, `deriv()`, and `deriv3()`.


## `D()` — symbolic derivatives

`D()` differentiates an R **expression** with respect to a named variable and returns a new expression.

::: {.small}
```{r}
D(expression(x^2), "x")
D(expression(sin(x) * cos(y)), "x")
D(expression(sin(x) * cos(y)), "y")
```
:::

. . .

::: {.small}
```{r}
D(expression(x^2 + 3*x*y + y^2), "x")
D(expression(x^2 + 3*x*y + y^2), "y")
```
:::


## Using `D()` to build gradient functions

We can evaluate the result of `D()` by providing variable values,

::: {.small}
```{r}
df_dx = D(expression(x^4 + x^3 - x^2 - x), "x")
df_dx
```
:::

. . .

::: {.small}
```{r}
eval(df_dx, list(x = 0))
eval(df_dx, list(x = 1))
```
:::

. . .

::: {.small}
```{r}
x = seq(-1.5, 1.5, length.out = 201)
plot(x, x^4 + x^3 - x^2 - x, type = "l", ylab = "f(x)")
lines(x, eval(df_dx, list(x = x)), col = "red", lty = 2)
abline(h = 0, col = "grey")
legend("top", c("f(x)", "f'(x)"), col = c("black", "red"), lty = c(1, 2))
```
:::


## Higher-order and chained derivatives

`D()` can be applied repeatedly for higher-order derivatives,

::: {.small}
```{r}
expr = expression(x^4 + x^3 - x^2 - x)
D(expr, "x")
D(D(expr, "x"), "x")
D(D(D(expr, "x"), "x"), "x")
```
:::

## `D()` limitations

`D()` only differentiates with respect to **one variable at a time** and returns an expression (not a function).

For multivariate functions where we want a callable function that returns both the value *and* gradient, we need `deriv()`.


## `deriv()` — gradient functions

`deriv()` takes a formula or expression and a character vector of variable names, and returns a function that computes the value and its `"gradient"` attribute.

::: {.small}
```{r}
f_grad = deriv(
  ~ x^2 + 3*x*y + y^2,
  namevec = c("x", "y"),
  function.arg = TRUE
)
f_grad
```
:::

. . .

::: {.small}
```{r}
f_grad(x = 1, y = 2)
```
:::


## Extracting the gradient

::: {.small}
```{r}
res = f_grad(x = 1, y = 2)
```
:::

. . .

::: {.small}
```{r}
drop(res)
attr(res, "gradient")
```
:::

. . .

The gradient at $(1, 2)$ is $(2(1) + 3(2),\; 3(1) + 2(2)) = (8, 7)$ — matching the analytical result.


## `deriv3()` — gradient *and* Hessian

`deriv3()` works exactly like `deriv()` but also computes the Hessian matrix (second derivatives), stored in a `"hessian"` attribute.

::: {.small}
```{r}
f_all = deriv3(
  ~ x^2 + 3*x*y + y^2,
  namevec = c("x", "y"),
  function.arg = TRUE
)
res = f_all(x = 1, y = 2)
drop(res)
attr(res, "gradient")
attr(res, "hessian")
```
:::


## Using `deriv()` with `optim()`

We can use `deriv()` to automatically supply gradients to `optim()`,

::: {.small}
```{r}
rosenbrock = deriv(
  ~ (1 - x)^2 + 100 * (y - x^2)^2,
  namevec = c("x", "y"),
  function.arg = TRUE
)
```
:::

. . .

::: {.small}
```{r}
fn = function(par) {
  res = rosenbrock(x = par[1], y = par[2])
  drop(res)
}
gr = function(par) {
  res = rosenbrock(x = par[1], y = par[2])
  drop(attr(res, "gradient"))
}
optim(c(-1, 1), fn, gr, method = "BFGS")
```
:::


## What `D()` / `deriv()` support

These functions handle a specific set of R math operations:

::: {.small}
| Supported | Examples |
|---|---|
| Arithmetic | `+`, `-`, `*`, `/`, `^` |
| Trig | `sin`, `cos`, `tan` |
| Logarithms | `log`, `exp`, `log2`, `log10` |
| Other math | `sqrt`, `abs`, `gamma`, `lgamma`, `beta`, `lbeta`, `digamma`, `trigamma` |
| Distribution | `dnorm`, `pnorm`, `dbinom`, `dpois`, `dgamma`, `dbeta`, `dt`, `df`, `dchisq`, `dweibull`, `dlnorm`, `dcauchy` |
:::

. . .

Not supported: control flow (`if`, `for`), matrix operations, or arbitrary R functions — for those you need numerical differentiation (e.g. `numDeriv::grad()`) or AD packages.

